---
title: "Threshold selection tools"
author: "Léo Belzile"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Threshold selection tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mev_vignette.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


The generalized Pareto (GP) distribution with scale $\sigma \in \mathbb{R}_{+}$ and shape $\xi \in \mathbb{R}$ is 
 \begin{align*}
  G(x)  = \begin{cases}
1-\left\{1+\xi \left(\frac{x}{\sigma}\right)\right\}_{+}^{-1/\xi}, &  \xi \neq 0,
\\ 1-\exp \left(-{x}{\sigma}\right),&  \xi = 0.
\end{cases}
 \end{align*}
The range of the generalized Pareto distribution is $[0, -\sigma/\xi)$ if $\xi < 0$ and $\mathbb{R}_{+}$ otherwise. 

Why is the generalized Pareto distribution so central to peaks-over-threshold analysis? The conditional distribution of exceedances over a threshold $u < x^*$ converges to a generalized Pareto distribution as $u$ converges to the endpoint of the distribution,
\begin{align*}
\lim_{u \to x^*} \frac{1-F(xa_u+u)}{1-F(u)} = 1-H(x),
\end{align*}
where $H(x)$ is the distribution function of $\mathsf{GP}(1, \xi)$. 

If $X \sim \mathsf{GP}(\sigma, \xi)$, straightforward calculations show that $X-u \mid X>u \sim \mathsf{GP}(\sigma + \xi u, \xi)$ for any $u \in \mathbb{R}$ such that $\sigma+\xi u>0$, meaning that conditional exceedances above a threshold $u$ also follow a generalized Pareto distribution. This property is termed threshold-stability. 

The limiting distribution of threshold exceedances is generalized Pareto as $u \to x^*$ but, in practice, we must choose a finite threshold in order to have enough exceedances to draw inference.  Since the scaling constant $a_u$ is unknown, we have $X \mid X > u \stackrel{\cdot}{\sim} \mathsf{GP}(\sigma_u, \xi)$. The term $1-F(u)$ in the denominator is the fraction of points exceeding the threshold, which has a binomial distribution.

Threshold selection is subtle  and it is common to select a high percentile of the data, say the 
95\% value, as the threshold, even if this is asymptotically incorrect, as in this case $k/n \nrightarrow 0$ as $n \to \infty$. Most approaches for threshold selection rely on properties of the 
generalized Pareto distribution (moments, threshold-stability) to determine a region within which the asymptotic distribution fits the data well and the parameter estimates are stable. 

Below, we focus  on recent graphical selection tools for likelihood based inference; mixture models are reviewed in @Scarrott:2012 and are available in the package `evmix`.

## Threshold stability plots

Consider a sequence of ordered candidate thresholds $u_1 < \cdots < u_k$; one of the most widely used tools for threshold selection is the threshold-stability plots of @Davison:1990. These show the point estimates of the shape $\xi$ and the modified scale $\sigma_{u_i}-\xi u_i$, which should be constant for any threshold $u_{j} >u_i$ assuming that the generalized Pareto above $u_i$ holds exactly. In addition to the point estimates, the asymptotic pointwise 95\% Wald confidence intervals are displayed; the standard errors are obtained from the observed information matrix. While these are displayed by many packages, notably `extRemes`, `ismev` and `evd` , `mev` allows you to use profile likelihood based confidence intervals, which typically offer better coverage and capture some of the asymmetry of the distribution. The `mev` package functions for producing threshold stability plots are `tstab.gpd` and `tstab.pp` for respectively the generalized Pareto and Poisson process likelihoods.

Parameter stability plots can be difficult to interpret because the confidence intervals are pointwise rather than simultaneous (each fit likewise uses an overlapping portion of the data). The plots also ignore changes in the estimated parameters due to the penultimate approximation. 

## Wadsworth's diagnostics: white noise process and simultaneous threshold stability plots

The problem with the threshold stability plots lies in the point-wise nature of the estimate. Assuming a superposition of $k$ Poisson processes, @Wadsworth:2016 derives the limiting distribution of the maximum likelihood estimators from the Poisson process for overlapping windows as the number of windows $k \to \infty$ and $n_k \to \infty$.  The joint asymptotic Gaussian distribution allows @Wadsworth:2016 to propose two additional diagnostics: a white noise sequence of differences in estimates of the shape, standardized to have unit variance. The variables $\xi^*_i=(\hat{\xi}_{u_{i+1}}-\hat{\xi}_{u_i})/\{(I^{-1}_{u_{i+1}}-I^{-1}_{u_{i}})_{\xi,\xi}^{1/2}\}$, where $I_{u_{i}}$ is the Fisher information of the Poisson process likelihood for exceedances above $u_i$, should form a white-noise sequence of independent variables centered around the origin; systematic deviations are indicative of inadequacy. To formally test the hypothesis, a likelihood ratio test can be used assuming a simple alternative, namely a single change point at threshold $u_j$. The null hypothesis is $\mathrm{H}_0: \xi_i^* \stackrel{\cdot}{\sim} \mathsf{No}(0,1)$ for $i=1, \ldots, k-1$  against the alternative  $\mathrm{H}_a: \xi^*_i \sim \mathsf{No}(\beta, \sigma) (i=1, \ldots, j-1)$ and $\xi^*_i \sim \mathsf{No}(0,1)$ for $j, \ldots, k-1$. This alternative is motivated by results on model misspecification [@White:1982], which suggest that the asymptotic distribution may still be Gaussian, but with a different mean and variance. This can be used to automate threshold selection, by picking the smallest threshold for which the $P$-value is above the level $\alpha$. The function `W.diag` returns diagnostics plots (for the likelihood ratio statistic path, the white noise process and threshold stability along with asymptotic simultaneous confidence intervals) for non-homogeneous Poisson process model and for the bivariate exponential and the  over a sequence of thresholds, specified using `q1`, `q2` and `k`. The argument `M` is a tuning parameter that can be chosen in a way such that the parameters of the non-homogeneous Poisson process likelihood coincide with those of the generalized extreme value distribution for blocks of size $m$; see @Coles:2001 to this effect. 

A main criticism of the proposals of @Wadsworth:2016 is their lack of robustness. For the asymptotic result to be approximately valid,  the number of thresholds must be large, which implicitly requires large samples for each superposed point process. Moreover, the estimated difference in Fisher information matrices often fails to be positive definite in practice. The procedure is highly sensitive to the choice of $k$. Changing the set of thresholds $\boldsymbol{u}$ under consideration leads to potentially completely different parameter estimates being chosen by the automated procedure.  


## Changepoint tests based on penultimate approximation

Let $F(x)$ denote a thrice-differentiable distribution function with endpoint $x^*$ and density $f(x)$. Define the reciprocal hazard function  $r(x) = \{1-F(x)\}/f(x)$. 
The existence of the limit $\xi_{\infty} = \lim_{n \to \infty} s'\{b_n\}$ is  necessary and sufficient for convergence to an extreme value distribution and @Smith:1987 shows that there exists $y$ such that
\begin{align*}
 \frac{1-F\{u+xr(u)\}}{1-F(u)} = \left\{1+xr'(y)\right\}_{+}^{-1/r'(y)}, \qquad u < y < u+xr(u),
\end{align*}
unless $r'(x)$ is constant. The penultimate shape parameter for the generalized Pareto distribution is $r'(u)$, but the true shape parameter lies between $r'(u)$ and $\xi_{\infty}$.

When we fit the limiting parametric models to finite samples, maximum likelihood estimates of the shape parameter will be closer to their penultimate counterparts than to the limiting value and we can expect them to vary as we increase the threshold.

@Northrop:2014 adapt the idea of @Wadsworth:2012 and fit a generalized Pareto model with piecewise constant shape to $k$ different thresholds; continuity constraints at the thresholds impose $k-1$ restrictions on scale parameters, so  the model only has $k+1$ parameters. A score test can be used to test the hypothesis of equal shape and it only requires evaluation of the model under the null hypothesis that a generalized Pareto distribution is valid above all thresholds. A diagnostic plot is obtained by plotting $P$-values against threshold. One can then choose to take, e.g., (a) the lowest threshold at which the $P$-value is non-significant, or (b) the lowest threshold at which the $P$-values for all higher thresholds are non-significant: under the null hypothesis, there is an $\alpha\%$ probability of rejection at any given threshold. The function `NC.diag` computes the $P$-value of the score test as a function of the threshold.


## Extended generalized Pareto

@Papastathopoulos:2013 propose three extended generalized Pareto distributions: for example, the third extended generalized Pareto model has distribution function $\{1-(1+\xi x/\sigma)^{-1/\xi}_{+}\}^{\kappa}$ for $x >0$ and $\kappa > 0$. Each family reduces to the generalized Pareto when the additional parameter $\kappa=1$ and share the same tail index $\xi$, the extended generalized Pareto provide more flexibility for modelling departures from the limiting form. Standard parameter stability plots  can be used to find a region in which $\kappa \approx 1$ and the shape parameter stabilizes. The additional parameter gives flexibility for modelling departures from the limiting distribution and the hope is one can fit to exceedances over lower threshold and increase the number of points to which the distribution is fitted. The stability plots, obtained through `tstab.egp`, suffer from the same caveats as classical diagnostics.

## Robust selection 

The extreme value distributions have unbounded influence functions and outliers can strongly affect the estimate of the shape. @Dupuis:1999 proposes an optimal $B$-robust estimator of the generalized Pareto parameters. Points that are outlying or for which the fit is poor are downweighted; if the weights for the largest observations are very low, this suggests that the threshold is too low. While there is no guarantee that observations that were simulated from a generalized Pareto distributed would not be downweighted, systematic downweighting of the largest exceedances may be indicative of poor fit. 

## Examples 

We are now ready to go back to the Maiquetia data and try to select a reasonable threshold. To this effect, we select a sequence of candidate thresholds and run the procedures described above. 

```{r tstabplots, out.width = '90%', fig.width = 8, fig.height = 10, fig.align = "center"}
library(mev)
data(maiquetia, package = "mev")
day <- seq.Date(from = as.Date("1961-01-01"), 
                to = as.Date("1999-12-31"), 
                by = "day")
nzrain <- maiquetia[substr(day, 1, 4) < 1999 & maiquetia > 0]
# Keep non-zero rainfall, exclude 1999 observations
nzrain <- maiquetia[substr(day, 3, 4) < 99 & maiquetia > 0]
thcan <- quantile(nzrain, 
                  seq(0.8, 0.99, length.out = 25))
tstab.gpd(dat = nzrain, 
          thresh = thcan, 
          method = "profile")
```

Another option is the @Northrop:2014 penultimate test, which considers a sequence of threshold and tests whether a single shape fits the data against the alternative where the shape parameter differs at each candidate threshold following a step function. The `NC.diag` function plots the $p$-value path against the candidate thresholds; the top axis also displays the number of threshold exceedances. The extended generalized Pareto distribution can also be used to produce threshold stability plot with 95% confidence intervals. The parameter $\kappa=1$ if the model is generalized Pareto. Although there are three potential families, the family `egp2` seems prefered for most settings.

```{r NCdiag, out.width = '90%',fig.width = 10, fig.height = 6, fig.align = "center", out.width = '90%'}
fNCdiag <- NC.diag(x = nzrain, u = thcan)
```


```{r tstagegp, out.width = '90%', fig.width = 8, fig.height = 15, fig.align = "center"}
tstab.egp(xdat = nzrain, 
          thresh = thcan, 
          model = "egp2")
```

What do you notice on the parameter stability plots? 

We can plot simultaneous parameter stability plots for the shape, as well as the white-noise sequence. The `W.diag` function, which implements the tests of @Wadsworth:2016, returns the white noise process and calculates by simulation the null distribution, selecting the lowest threshold at which the hypothesis of changepoint is not rejected for all higher thresholds. For the Maiquetia rainfall, a threshold around 15.15mm is returned.

```{r Wdiag, out.width = '90%',fig.width = 10, fig.height = 12, fig.align = "center"}
fWdiag <-
  W.diag(
    xdat = nzrain,
    model = "nhpp",
    u = thcan,
    plots = c("WN", "PS")
  )
```


## Your turn

1. Try using the threshold diagnostic tests on the `nidd` river flow data. Which threshold would you choose based on the conclusions of the test? Repeat the analysis on your favorite dataset. 
2. Compare the inference obtained by using the different threshold (by comparing, for example, return levels or quantiles of a distribution of interest). Are your results sensitive to the choice of threshold?

## References
